Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2015-01-31 draft_lm1.tex
An important (caveat) note
small tweaks

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Estimating the Attack ratio of Dengue Epidemics, under time-varying 
force of infection, from aggregated case notification data}
}
\\
Flavio Code\c{c}o Coelho$^{1\ast}$, 
Luiz Max de Carvalho$^{2}$, 
\\
\bf{1} Escola de Matem\'atica Aplicada , Funda\c{c}\~ao Getulio Vargas (FGV), 
Rio de Janeiro -- RJ, Brazil.
\\
\bf{2} Programa de Computa\c{c}\~ao Cient\'ifica (PROCC), Funda\c{c}\~ao Oswaldo 
Cruz, Rio de Janeiro -- RJ, Brazil.
\\
$\ast$ E-mail: fccoelho@fgv.br
\end{flushleft}

@@

\section*{Author Summary}

@@

The Attack ratio (AR) of a disease is a measure of morbidity defined as the 
number of new cases divided by the population at risk.
For Dengue outbreaks and epidemics, it can be difficult to calculate the AR due 
to the lack of knowledge of the population at risk. The population at risk 
in the case of dengue, is the number of susceptibles to the circulating virus 
type(s) before a given epidemic.

@@

Many authors have chosen to quantify the uncertainty about $\theta$ 
following orthodox approaches  (see for example~\cite{wilson} 
and~\cite{clopper}) mainly for simplicity.
We choose instead to take a Bayesian approach and use the  $100\alpha \%$ 
posterior credibility interval for $\theta_t$ as $c_{\alpha}(\theta_t)$.
If we choose the conjugate beta prior with parameters $a_0$ and $b_0$ for the 
binomial likelihood in (\ref{eq:binlike}), the posterior distribution for 
$\theta_t$ is
\begin{equation}
\label{eq:thetapost}
p(\theta_t| Y_{t+1}, S) \sim Beta(Y_{t+1} + a_0, Y_t + b_0)
\end{equation}
Combining equations~(\ref{eq:confRt}) and~(\ref{eq:thetapost}) 
tells us that the induced posterior distribution of $R_t$ is 
a beta prime (or inverted beta) with parameters $ a_1 = Y_{t+1} + a_0$ and $b_1 
=  Y_t + b_0$~\cite{dubey1970}.
The density of the induced distribution is then 
\begin{equation}
\label{eq:densityMantel}
f_P(R_t| a_1, b_1) = \frac{\Gamma(a_1 + b_1)}{\Gamma(a_1)\Gamma(b_1)} R_t^{a_1 - 
1} (1 + R_t)^{-(a_1 + b_1)}
\end{equation}
Thus, the expectation of \rr is $a_1/(b_1 - 1)$ and its variance is 
$a_1(a_1 + b_1 - 1)/\left((b_1 - 2)(b_1 - 1)^2 \right) $.
Sampling from the posterior in~(\ref{eq:densityMantel}) can be made 
straightforward by first sampling from~(\ref{eq:thetapost}) and then applying 
the transform in~(\ref{eq:confRt}).
Also, one can choose $a_0$ and $b_0$ so as to elicit meaningful prior 
distributions for $R_t$.
We show how to elicit the prior for $R_t$ from specified prior mean and variance
or coefficient of variation in the Appendix.

@@

\begin{table}[!ht]
\caption{
{\bf Median Attack ratio and 95\% credibility intervals calculated according to 
(\ref{eq:AR2})}. Values are presented as percentage of total population. 
$^\dag$: Year corresponds to the start of the epidemic. 
Sometimes the peak of cases occur in the following year.
$^\ddag$: Susceptible fraction.}
\begin{center}
\begin{tabular}{c|c|c}
\hline
Year$^\dag$ & Median attack ratio (\%) & $S_0^\ddag$ (\%)\\
\hline
1996 & 38.64 (17.18-54.19) & 0.171\\
1997 & 87.31 (73.83-87.31) & 0.273\\
1998 & 49.86 (49.19-49.86) & 0.142\\
1999 & 10.76 (3.72-20.24) & 0.345\\
2000 & 25.49 (24.47-26.98) & 1.547\\
2001 & 48.12 (46.57-49.15) & 4.951\\
2005 & 14.74 (10.23-21.38) & 1.472\\
2006 & 11.22 (8.00-14.41) & 2.810\\
2007 & 15.08 (14.76-15.38) & 13.451\\
2008 & 13.73 (3.10-30.84) & 0.672\\
2010 & 18.30 (17.34-19.33) & 4.540\\
2011 & 8.59 (8.20-9.36) & 21.482\\
2012 & 13.92 (12.71-14.92) & 6.208\\

>>>>

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Estimating the attack ratio of Dengue Epidemics, under time-varying 
force of infection, from aggregated case notification data}
}
\\
Flavio Code\c{c}o Coelho$^{1\ast}$, 
Luiz Max de Carvalho$^{2}$, 
\\
\bf{1} Escola de Matem\'atica Aplicada , Funda\c{c}\~ao Getulio Vargas (FGV), 
Rio de Janeiro -- RJ, Brazil.
\\
\bf{2} Programa de Computa\c{c}\~ao Cient\'ifica (PROCC), Funda\c{c}\~ao Oswaldo 
Cruz, Rio de Janeiro -- RJ, Brazil.
\\
$\ast$ E-mail: fccoelho@fgv.br
\end{flushleft}

@@

@@

The attack ratio (AR) of a disease is a measure of morbidity defined as the 
number of new cases divided by the population at risk.
For Dengue outbreaks and epidemics, it can be difficult to calculate the AR due 
to the lack of knowledge of the population at risk. The population at risk 
in the case of dengue, is the number of susceptibles to the circulating virus 
type(s) before a given epidemic.

@@

Many authors have chosen to quantify the uncertainty about $\theta$ 
following orthodox approaches  (see for example~\cite{wilson} 
and~\cite{clopper}) mainly for simplicity.
We choose instead to take a Bayesian approach and use the  $100\alpha \%$ 
posterior credibility interval for $\theta_t$ as $c_{\alpha}(\theta_t)$.
If we choose the conjugate beta prior with parameters $a_0$ and $b_0$ for the 
binomial likelihood in (\ref{eq:binlike}), the posterior distribution for 
$\theta_t$ is
\begin{equation}
\label{eq:thetapost}
p(\theta_t| Y_{t+1}, S) \sim Beta(Y_{t+1} + a_0, Y_t + b_0)
\end{equation}
Combining equations~(\ref{eq:confRt}) and~(\ref{eq:thetapost}) 
tells us that the induced posterior distribution of $R_t$ is 
a beta prime (or inverted beta) with parameters $ a_1 = Y_{t+1} + a_0$ and $b_1 
=  Y_t + b_0$~\cite{dubey1970}.
The density of the induced distribution is then 
\begin{equation}
\label{eq:densityMantel}
f_P(R_t| a_1, b_1) = \frac{\Gamma(a_1 + b_1)}{\Gamma(a_1)\Gamma(b_1)} R_t^{a_1 - 
1} (1 + R_t)^{-(a_1 + b_1)}
\end{equation}
Thus, the expectation of \rr is $a_1/(b_1 - 1)$ and its variance is 
$a_1(a_1 + b_1 - 1)/\left((b_1 - 2)(b_1 - 1)^2 \right) $.
Note that this result holds only for $n = 1$.
Sampling from the posterior in~(\ref{eq:densityMantel}) can be made 
straightforward by first sampling from~(\ref{eq:thetapost}) and then applying 
the transform in~(\ref{eq:confRt}).
Also, one can choose $a_0$ and $b_0$ so as to elicit meaningful prior 
distributions for $R_t$.
We show how to elicit the prior for $R_t$ from specified prior mean and variance
or coefficient of variation in the Appendix.

@@

\begin{table}[!ht]
\caption{
{\bf Median attack ratio and 95\% credibility intervals calculated according to 
(\ref{eq:AR2})}. Values are presented as percentage of total population. 
$^\dag$: Year corresponds to the start of the epidemic. 
Sometimes the peak of cases occur in the following year.
$^\ddag$: Susceptible fraction.}
\begin{center}
\begin{tabular}{c|c|c}
\hline
Year$^\dag$ & Median attack ratio (\%) & $S_0^\ddag$ (\%)\\
\hline
1996 & 38.64 (17.18-54.19) & 0.171\\
1997 & 87.31 (73.83-87.31) & 0.273\\
1998 & 49.86 (49.19-49.86) & 0.142\\
1999 & 10.76 (3.72-20.24) & 0.345\\
2000 & 25.49 (24.47-26.98) & 1.547\\
2001 & 48.12 (46.57-49.15) & 4.951\\
2005 & 14.74 (10.23-21.38) & 1.472\\
2006 & 11.22 (8.00-14.41) & 2.810\\
2007 & 15.08 (14.76-15.38) & 13.451\\
2008 & 13.73 (3.10-30.84) & 0.672\\
2010 & 18.30 (17.34-19.33) & 4.540\\
2011 & 8.59 (8.20-9.36) & 21.482\\
2012 & 13.92 (12.71-14.92) & 6.208\\
